Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat

### 2023-11-07 manuscript.Rmd
various updates

library(tidyverse)
library(pander)
library(fs)
library(here)
library(english)
library(phyloseq) 
library(vegan)
library(ggrepel)
library(here)
library(worrms)
library(BiodiversityR)
library(ggtext)
```

@@

ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol. Specimens were collected under permits [#xxx]{.mark} (NWHI) and [#xxx]{.mark} (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).

@@

Potential environmental drivers of shallow crab community structure were assembled from a study on factors affecting population genetics in Hawai&#x02BB;i, [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]), NASA Physical Oceanograpy Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC). For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental meterial. Due to logistical limitations, environmental factors were collected only for shallow samples ($\le$ 30 m).

@@

For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}A]; `r k_test(all_alpha$k_r_island)`), but Simpson diversity was significantly different ([@fig:fig_alpha[]{}C]; `r k_test(all_alpha$k_s_island,plain=TRUE)`). Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}D]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}A]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}B]).

<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}A]; `r w_test(all_alpha$w_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`). Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).  -->

<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}A]; `r t_test(all_alpha$t_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`). Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).  -->

@@

```{r rda-analysis, results="hide", dependson="init"}
library(formula.tools)

@@

Environmental variables clustered strongly according to island group (MHI vs NWHI), with the difference driven primarily by human impact, coral cover, and island slope ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated. The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).

```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands. Ellipses represent 95% confidence intervals around groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
termz <- model_terms %>%
  select(variable,display=`Environmental Driver`)

@@

# run CAP analysis, taking care of negative eigenvalues where necessary
cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
# # calculate proportion of explained variance
cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
cap_var <- list(x = cap_var[1], y = cap_var[2])
```
<!-- 
Here's the thing: it's not that the differences between MHI & NWHI become non-significant with the inclusion of deep samples, 
it's just that the deep samples are MORE different from all the shallow samples than all the shallow samples are from one another?
We can think of PERMANOVA R^2 and ANOSIM R as effect sizes. Thus, it may be a bit weak but for shallow samples only, island group 
R2 = 0.072, for all samples R2 = 0.065. For depth (all samples of course), R2 = 0.110. If we consider the effect sizes from ANOSIM, 
for shallow samples, island group R = 0.175, for all samples, island group R = 0.161, and for shallow vs deep,  R = 0.509. 
The apparent difference in effect size is much more pronounced in the ANOSIM, but it's different in each case, and it's smaller
for all samples than it is for just shallow samples.
-->

Shallow crab community structure was significantly different between island groups (MHI vs. NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons). There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`). When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`). However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`). Although crab communities varied significantly among both depth zones and island groups when examining the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$). In the CAP analysis, ARMS units were correctly assigned to sampling groups `r pc(cap_depth$percent/100,0.1)` of the time. The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively. The mesophotic samples ($\geq$ 30 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]). In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread along the LD2 (vertical) axis and the 60 m and 90 m samples grouped together ([@fig:fig_cap]).

@@

# Discussion {-}

@@

@@

pcdata <- cds %>%
  select(where(is.numeric),-c(lon,lat))

@@

```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
library(ggtext)
pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=PERM)

>>>>

library(tidyverse)
library(pander)
library(fs)
library(here)
library(english)
library(phyloseq) 
library(vegan)
library(ggrepel)
library(here)
library(worrms)
library(BiodiversityR)
library(ggtext)

# set random seed for consistency
set.seed(31337)
```

@@

ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol. Specimens were collected under permits [#xxx]{.mark} (NWHI) and [#xxx]{.mark} (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).

@@

Potential environmental drivers of shallow crab community structure were assembled a previous study on environmental drivers of genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), NASA Physical Oceanography Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC) ([@tbl:env_table]). For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental material. Due to logistical limitations, environmental data were collected only for shallow samples ($\le$ 30 m).

@@

For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`). Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).

<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`). Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).  -->

<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r t_test(all_alpha$t_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`). Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).  -->

@@

```{r dbrda-analysis, results="hide", dependson="init"}
library(formula.tools)

@@

Environmental variables were strongly clustered according to island group (MHI vs NWHI), with the separation primarily along the direction influenced by human impact, coral cover, sea-surface temperature, and chlorophyll ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the seven environmental variables tested were significant and sufficiently uncorrelated. The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).

```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands. Colored ellipses represent 95% confidence intervals around island groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
termz <- model_terms %>%
  select(variable,display=`Environmental Driver`)

@@

# run CAP analysis, taking care of negative eigenvalues where necessary
cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
# # calculate proportion of explained variance
cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
cap_var <- list(x = cap_var[1], y = cap_var[2])
```
Shallow crab community structure was significantly different between island groups (MHI vs. NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons). There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`). When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`). However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`). Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).

In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to groups `r pc(cap_depth$percent/100,0.1)` of the time. The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively. Mesophotic samples ($\geq$ 30 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]). Additionally, variability among 30--90 m samples was equivalent to that among all shallow samples, although that was largely due to differences between 30 m and deeper sample groups ([@fig:fig_cap]).

@@

# Discussion {-}
The Hawaiian Archipelago is among the most remote on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal area to study the impacts of these factors on species assemblages across space. Human impacts are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016]. Although we did not consider individual effects here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to previous work examining genetic diversity across the archipelago [@selkoeDNACoralReef2016]. The next-most significant driver, chlorophyll a, can be used as a proxy for primary productivity. Productivity gradients have been reported across the Hawaiian Archipelago, which coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], and survival rates of top predators [@bakerEffectVariableOceanic2007]. It is not surprising, therefore, that we find differences in the brachyuran crab communities are driven by these same factors across the Hawaiian archipelago.

@@

## Processing of satellite-derived SST and chl-A data
Sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @NASAJPL_2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively. Monthly summary datasets were downloaded for 30 randomly subsampled months within each ARMs unit's three-year soaking period, averaged, and values (SST or chl-A) were extracted for the geographic coordinates of each sampling site, with a 300 m surrounding buffer zone.

@@

pcdata <- cds %>%
  select(where(is.numeric),-c(lon,lat),-closest_island)

@@

```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate [@benjaminiControllingFalseDiscovery1995]).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
library(ggtext)
pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=PERM)
